Graph Topological Aspects of Granger Causal Network Learning

### 2019-08-27 paper_appl/paper_appl.tex
some preliminary analysis on GC appl paper.

>>>>

\section{Laten Structure Recovery Performance Metric}
The Granger causal graph structure described in section
\ref{sec:theory} is entirely latent.  That is, we can never observe
the true causality graph (assuming such a structure \textit{exists}),
which makes evaluating different support recovery algorithms
difficult.  Broadly speaking, most work on the topic evaluates the
performance of competing causality graph recovery algorithms in one of
two ways (or both): (1) theoretical proofs or (2) simulations where
the true graph is known by construction.  In this work our goal is to
establish as valid a third possibility: the performance of a second
step classifier.

\subsection{Introduction and Motivation}
More precisely, we suppose we have a two distinct data generating
processes (``subjects'') labeled $S^{(1)}$ and $S^{(2)}$.  Each
subject has their own fixed and unobservable Granger causality graph
$\gcg^{(i)}$.  And, when queried, the subject returns a finite sample
of data $\{x^{(i)}(t)\}_{t = 1}^{T}$ generated by the model
\ref{eq:var_model}.  In addition, we have $K$ algorithms
$A_1, \ldots, A_K$ which consume such a dataset and return an estimate
$\widehat{\gcg}_k^{(i)}$ of graph $\gcg^{(i)}$.

Following this process, we suppose that we collect $N^{(i)}$ samples
from each subject.  Applying each algorithm to the resulting time
series data results in $NK$ datasets

\[
  \mathcal{D}_k = \bigcup_{i \in \{0, 1\}} \bigcup_{n = 1}^{N^{(i)}} \{(\widehat{\gcg}_{k, n}^{(i)}, i)\}.
\]

Finally, we fit classifiers $C_k: \widehat{\gcg} \rightarrow \{0, 1\}$
which consume an estimated causality graph and attempt to return the
subject label $i \in \{0, 1\}$ that generated the data from which the
graph was estimated.  We will conclude that Algorithm $k_1$ is more
accurate at estimating causality graphs than Algorithm $k_2$ if

\begin{equation}
  \label{eqn:comparison_metric}
  \P\big\{C_{k_1}(\widehat{\gcg}_{k_1}^{(i)}) = i\big\} > \P\big\{C_{k_2}(\widehat{\gcg}_{k_2}^{(i)}) = i\big\}.
\end{equation}

That is, if it is easier to label the subject from causality graphs
estimated by $A_{k_1}$ than it is to do so via $A_{k_2}$, it provides
evidence that $\widehat{\gcg}_{k_1}^{(i)}$ is closer to
$\gcg^{(i)}$ than is $\widehat{\gcg}_{k_2}^{(i)}$.

\begin{remark}
We will provide mathematical justification for this evaluation
strategy below, but some remarks are in order.  Firstly, if Equation
\ref{eqn:comparison_metric} holds, it does not constitute
\textit{proof} that $A_{k_1}$ is superior to $A_{k_2}$ for the task of
estimating Granger causality graphs.  Indeed, it would be possible to
construct an algorithm $A$ which attempts to encode the idiosyncracies
of $s^{(i)}$ present in the data $x^{(i)}(t)$, making no attempt at
actually recovering $\gcg^{(i)}$.  We therefore formulate our
theoretical justification such that we make no reference to the actual
recovery algorithm.  Secondly, we claim that this evaluation strategy
is distinct from the above evaluation metrics (1) and (2), but that we
do not claim it is \textit{better} than simulation or mathematical
argument.  Merely, we propose it as a third mechanism for comparing
different recovery algorithms that may provide further insight beyond
(1) and (2).
\end{remark}

\subsection{Theoretical Analysis}

### 2019-08-27 paper_theory/graph_topology_granger_causality.tex
some preliminary analysis on GC appl paper.

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\MA{\mathsf{MA}}  % MA(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

@@

\abstract{We study Granger-causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO-type algorithms) when structural assumptions are met.}

\keywords{causality graph, gene regulatory networks,
  Granger-causality, network learning, time series, vector
  autoregression}

@@

\section{Introduction and Review}
\label{sec:introduction}
In this paper we study the notion of Granger-causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly,
it's presence is inferred as a latent structure among observed
time series data.  This notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

Granger-causality can generally be formulated by searching for the
``best'' graph structure consistent with observed data, which is in
general an extremely challenging problem (i.e. it may be framed as a
best subset selection problem, see, \cite{hastie_bss_comp} for recent
improvements in BSS methods), moreover, the comparison of quality
between different structures, and hence the notion of ``best'' needs
qualification.  In applications where we are interested merely in
minimizing the mean squared error of a linear one-step-ahead
predictor, then we will be satisfied with an entirely dense graph of
connections, since each edge can only serve to reduce estimation
error.  However, since the number of edges scales quadratically in $n$
(the number of nodes) it becomes imperative to infer a sparse
causality graph for large systems, both to avoid overfitting observed
data, as well as to aid the interpretability of the results.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger-causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.

% We study a novel perspective on the problem by establishing conditions
% under which simple \textit{pairwise} Granger-causality tests are
% sufficient to recover the entire graph.  Without particular
% topological assumptions, pairwise testing naturally results in a large
% number of false positives \cite{tam2013gene_pwgc}.

In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger-causality graph.  The focus of this paper is to shed light on
some of these topological questions, in particular, we study a
particularly simple notion of graph topology which we term
``strongly causal'' and show that stationary times series whose
underlying causality graph has this structure satisfy natural
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly recoverable with only
\textit{pairwise} Granger-causality tests, which would otherwise
suffer from serious confounding problems (see
e.g. \cite{tam2013gene_pwgc} for earlier work on pairwise testing).
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

@@

The principle contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger-causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger-causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger-causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report a summary of simulation results in \ref{sec:simulation}, with
additional results reported in the supplementary material Section
\ref{apx:simulation}.  Our simulation results establish that there is
significant potential for improvement over existing methods, and that
the graph-topological aspects of time series analysis are relevant for
both theory and practice.  Concluding remarks on further open problems
and extensions are provided in Section \ref{sec:conclusion}.  The
proofs of each proposition and theorem are also relegated to the
supplementary material, simple corollaries have proofs included in the
main text.

\section{Graph Topological Aspects of Granger-causality}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.
We suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has
absolutely summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$, and an absolutely continuous
spectral density.

@@

The notion of conditional orthogonality is the essence of
Granger-causality, and enables us to obtain results for a fairly
general class of WSS processes, rather than simply $\VAR(p)$ models.

@@

Our principle object of study will be a graph determined by
Granger-causality relations as follows.

\begin{definition}[Causality graph]
  We define the Granger-causality graph $\gcg = ([n], \gcge)$ to be the directed
  graph formed on $n$ vertices where an edge $(j, i) \in \gcge$ if and
  only if $x_j$ Granger-causes $x_i$ (with respect to $x$).  That is,
  $$(j, i) \in \gcge \iff j \in \pa{i} \iff x_j \gc x_i.$$
\end{definition}

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with the
\textit{strictly causal} LTI filter $\B_{ij}(z)$ (see Equation
\eqref{eqn:filter_action}).  Thence, the matrix $\B(z)$ is analogous
to a \textit{weighted adjacency matrix}\footnote{\footnotesize We are using the
  convention that $\B_{ij}(z)$ is a filter with input $x_j$ and output
  $x_i$ so as to write the action of the system as $\B(z)x(t)$ with
  $x(t)$ as a column vector.  This competes with the usual convention
  for adjacency matrices where $A_{ij} = 1$ if there is an edge
  $(i, j)$.  In our case, the sparsity pattern of $\B_{ij}$ is the
  \textit{transposed} conventional adjacency matrix.} for the graph $\gcg$.  And,
in the same way that the $k^{\text{th}}$ power of an adjacency matrix
counts the number of paths of length $k$ between nodes,
$(\B(z)^k)_{ij}$ is a filter isolating the ``action'' of $j$ on $i$ at
a time lag of $k$ steps, this is exemplified in the inversion formula
\eqref{eqn:lsi_inversion}.

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed, completely
explains the connection from $x_j$ to $x_i$.  In this section we study
\textit{pairwise} Granger-causality, and seek to understand when
knowledge of pairwise relations is sufficient to deduce the true fully
conditional relations of $\gcg$.

\begin{definition}[Pairwise Granger-causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect only to
  $(x_i, x_j)$.
\end{definition}

@@

\begin{proposition}
  Consider distinct nodes $i, j$ in a Granger-causality graph
  $\gcg$.  If

@@

\begin{proposition}
  \label{prop:ancestor_properties}
  If in a Granger-causality graph $\gcg$ where $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

@@

\begin{definition}[Strongly Causal]
  \label{def:strongly_causal}
  We will say that a Granger-causality graph $\gcg$ is
  \textit{strongly causal} if there is at most one directed path between
  any two nodes.  Strongly Causal Graphs will be referred to as SCGs.
\end{definition}

@@

\begin{definition}[Persistent]
  We will say that the process $x(t)$ with Granger-causality graph
  $\gcg$ is \textit{persistent} if for every $i \in [n]$ and every
  $k \in \anc{i}$ we have $\tau_0(\A_{ik}) < \infty$ and $\tau_\infty(\A_{ik}) = \infty$.
\end{definition}

\begin{remark}
  In the context of Granger-causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, see Example
  \ref{ex:persistent_system}.

@@

\begin{theorem}[Pairwise Recovery]
  \label{thm:scg_recovery}
  If the Granger-causality graph $\gcg$ for the process $x(t)$ is a
  strongly causal DAG and Assumption \ref{ass:T_causality} holds, then
  $\gcg$ can be inferred from pairwise causality tests.  The procedure
  can be carried out, assuming we have an oracle for pairwise
  causality, via Algorithm (\ref{alg:pwgr}).
\end{theorem}

@@

  % \BlankLine
  \caption{Pairwise Granger Causality Algorithm}
  \label{alg:pwgr}
  \TitleOfAlgo{Pairwise Graph Recovery}
  \Input{Pairwise Granger-causality relations between a persistent
  process of dimension $n$ whose joint Granger-causality
  relations are known to form a strongly causal DAG $\gcg$.}
  % \Input{Pairwise Granger-causality relations}
  \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
    the graph $\gcg$.}
  \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
    $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
    % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
    $k = 1$ \texttt{\# a counter used only for notation}}
  \BlankLine
  $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
  $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parent-less nodes}\\
  \While{$S_{k - 1} \ne \emptyset$}{
    $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
    $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children}\\

@@

%   contains a loop but is pairwise recoverable, though not by algorithm
%   (\ref{alg:pwgr}).  Clearly, this example is somewhat artificial
%   since when $n = 2$ there is no difference between pairwise
%   Granger-causality and joint Granger-causality amongst all series --
%   however, one can add any number of nodes having no parents or
%   children to a graph containing a length 2 cycle, in which case the
%   graph clearly remains pairwise recoverable.
% \end{example}

@@

\begin{remark}[MCC as a Support Recovery Measurement]
  We apply Matthew's Correlation Coefficient (MCC)
  \cite{matthews1975comparison} as a statistic for measuring support
  recovery performance.  This statistic synthesizes the confusion
  matrix into a single score appropriate for unbalanced labels and is
  calibrated to fall into the range $[-1, 1]$ with $1$ being perfect
  performance, $0$ being the performance of random guessing, and $-1$
  being perfectly opposed.
\end{remark}

@@

\section{Conclusion}
\label{sec:conclusion}
In this paper we have argued that considering particular topological
properties of Granger-causality networks can provide substantial
insights into the structure of causality graphs with potential for
providing improvements to causality graph estimation when structural
assumptions are met.  In particular, the notion of a strongly-causal
graph has been exploited to establish conditions under which pairwise
causality testing alone is sufficient for recovering a complete
Granger-causality graph.  Moreover, examples from the literature
suggest that such topological assumptions may be reasonable in some
applications.  And secondly, even when the strong-causality assumption
is not met, we have provided simulation evidence to suggest that our
pairwise testing algorithm PWGC can still outperform the LASSO and
adaLASSO, both of which are commonly employed in applications.

We emphasize that the causality graph topology is one of the key
defining features of time series analysis in comparison to standard
multivariate regression and therefore advocate for further study of
how different topological assumptions may impact the recovery of
causality graphs.  For example, are there provable guarantees on the
error rate of PWGC when applied to non strongly-causal graphs?  Can
constraint systems or cunning adaptive weighting schemes impose useful
prior knowledge about graph topology for the LASSO algorithm?
Finally, the work of \cite{barnett2015granger} has established the
superiority of Granger-causality testing by state space models (as
opposed to pure autoregressions) in many cases.  Combining this work
with our PWGC algorithm (by modifying the approach described in
Section \ref{sec:pairwise_hypothesis_testing} to instead utilize
state-space Granger-causality testing) therefore is likely to enable
application to very large networks of time series data which are not
well approximated by finite $\VAR(p)$ models.  Moreover, our
heuristics are in principle applicable in a model-free context.  As
long as a primitive for testing the pairwise causation between two
components of a multivariate time series is available, our methods may
be useful.

>>>>

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\MA{\mathsf{MA}}  % MA(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

@@

\abstract{We study Granger causality in the context of wide-sense
  stationary time series, where our focus is on the topological
  aspects of the underlying causality graph.  We establish sufficient
  conditions (in particular, we develop the notion of a ``strongly
  causal'' graph topology) under which the true causality graph can be
  recovered via pairwise causality testing alone, and provide examples
  from the gene regulatory network literature suggesting that our
  concept of a strongly causal graph may be applicable to this field.
  We implement and detail finite-sample heuristics derived from our
  theory, and establish through simulation the efficiency gains (both
  statistical and computational) which can be obtained (in comparison
  to LASSO-type algorithms) when structural assumptions are met.}

\keywords{causality graph, Granger causality, network learning, time
  series, vector autoregression, LASSO}

@@

\section{Introduction and Review}
\label{sec:introduction}
In this paper we study the notion of Granger causality
\cite{granger1969investigating} \cite{Granger1980329} as a means of
uncovering an underlying causal structure in multivariate time series.
Though the underlying causality graph cannot be observed directly,
it's presence is inferred as a latent structure among observed
time series data.  This notion is leveraged in a variety of
applications e.g. in Neuroscience as a means of recovering
interactions amongst brain regions \cite{bressler2011wiener},
\cite{anna_paper2008}, \cite{david2008identifying}; in the study of
the dependence and connectedness of financial institutions
\cite{NBERw16223}; gene expression networks \cite{Fujita2007},
\cite{methods_for_inferring_gene_regulatory_networks_from_time_series_expression_data},
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery},
\cite{discovering_graphical_Granger_causality_using_the_truncating_lasso_penalty};
and power system design \cite{Misyrlis2016450}, \cite{yuan2014root}.

Granger causality can generally be formulated by searching for the
``best'' graph structure consistent with observed data, which is in
general an extremely challenging problem (i.e. it may be framed as a
best subset selection problem, see, \cite{hastie_bss_comp} for recent
improvements in BSS methods), moreover, the comparison of quality
between different structures, and hence the notion of ``best'' needs
qualification.  In applications where we are interested merely in
minimizing the mean squared error of a linear one-step-ahead
predictor, then we will be satisfied with an entirely dense graph of
connections, since each edge can only serve to reduce estimation
error.  However, since the number of edges scales quadratically in $n$
(the number of nodes) it becomes imperative to infer a sparse
causality graph for large systems, both to avoid overfitting observed
data, as well as to aid the interpretability of the results.

@@

A second and wildly successful heuristic is the LASSO regularizer
\cite{tibshirani1996regression}, which can be understood as a natural
convex relaxation to penalizing the count of the non-zero edges.  The
LASSO enjoys fairly strong theoretical guarantees
\cite{wainwright2009sharp}, extending largely to the case of
stationary time series data with a sufficiently fast rate of
dependence decay \cite{basu2015} \cite{wong2016lasso}
\cite{autoregressive_process_modelling_via_the_lasso_procedure}, and
variations on the LASSO have been applied in a number of different
time series contexts as well as Granger causality
\cite{DBLP:journals/corr/HallacPBL17} \cite{haufe2008sparse}
\cite{bolstad2011causal} \cite{he2013stationary}
\cite{grouped_graphical_granger_modelling_for_gene_expression_regulatory_networks_discovery}.
One of the key improvements to the original LASSO algorithm is the
adaptive (i.e. weighted) ``adaLASSO'' \cite{adaptive_lasso_zou2006},
for which oracle results (i.e. asymptotic support recovery) are
established under less restrictive conditions than for the vanilla
LASSO.

% We study a novel perspective on the problem by establishing conditions
% under which simple \textit{pairwise} Granger causality tests are
% sufficient to recover the entire graph.  Without particular
% topological assumptions, pairwise testing naturally results in a large
% number of false positives \cite{tam2013gene_pwgc}.

In the context of time series data, sparsity assumptions remain
important, but there is significant additional structure that may
arise as a result of considering the topology of the underlying
Granger causality graph.  The focus of this paper is to shed light on
some of these topological questions, in particular, we study a
particularly simple notion of graph topology which we term
``strongly causal'' and show that stationary times series whose
underlying causality graph has this structure satisfy natural
intuitive notions of ``information flow'' through the graph.
Moreover, we show that such graphs are perfectly recoverable with only
\textit{pairwise} Granger causality tests, which would otherwise
suffer from serious confounding problems (see
e.g. \cite{tam2013gene_pwgc} for earlier work on pairwise testing).
Aside from being an interesting theoretical perspective, prior
assumptions about the underlying graph (similarly to sparsity
assumptions) can greatly improve upon the statistical power of
causality graph recovery algorithms when the assumptions are met.

Detailed study of Granger causality for star structured graphs has
been carried out in \cite{jozsa2018relationship}.  See as well
\cite{barnett2015granger}, \cite{jozsa2018_phd_thesis} for state space
formulations.

@@

The principal contributions of this paper are as follows: firstly, in
section \ref{sec:theory} we study \textit{pairwise} Granger causality
relations, providing novel theorems connecting the structure of the
causality graph to the pairwise ``causality flow'' in the system, as
well as an interpretation in terms of the graph topology of the
sparsity pattern of matrices arising in the Wold decomposition,
generalizing in some sense the notion of ``feedback-free'' processes
studied by \cite{caines1975feedback} in close connection with
Granger causality.  We establish sufficient conditions (sections
\ref{sec:strongly_causal_graphs}, \ref{sec:persistent_systems}) under
which a fully conditional Granger causality graph can be recovered
from pairwise tests alone (sec \ref{sec:pairwise_algorithm}).  We
report a summary of simulation results in \ref{sec:simulation}, with
additional results reported in the supplementary material Section
\ref{apx:simulation}.  Our simulation results establish that there is
significant potential for improvement over existing methods, and that
the graph-topological aspects of time series analysis are relevant for
both theory and practice.  Concluding remarks on further open problems
and extensions are provided in Section \ref{sec:conclusion}.  The
proofs of each proposition and theorem are also relegated to the
supplementary material, simple corollaries have proofs included in the
main text.

\section{Graph Topological Aspects of Granger causality}
\label{sec:theory}
\subsection{Formal Setting}
Consider the space $L_2(\Omega)$, the usual Hilbert space of finite
variance random variables over a probability space
$(\Omega, \mathcal{F}, \mathbb{P})$ having inner product
$\inner{x}{y} = \E[xy]$.  We will work with a discrete time and
wide-sense stationary (WSS) $n$-dimensional vector valued process
$x(t)$ (with $t \in \Z$) where the $n$ elements take values in $L_2$.
We suppose that $x(t)$ has zero mean, $\E x(t) = 0$, and has
absolutely summable matrix valued covariance sequence
$R(\tau) \overset{\Delta}{=} \E x(t)x(t - \tau)^\T$, and an absolutely continuous
spectral density.

@@

The notion of conditional orthogonality is the essence of
Granger causality, and enables us to obtain results for a fairly
general class of WSS processes, rather than simply $\VAR(p)$ models.

@@

Our principal object of study will be a graph determined by
Granger causality relations as follows.

\begin{definition}[Causality graph]
  We define the Granger causality graph $\gcg = ([n], \gcge)$ to be the directed
  graph formed on $n$ vertices where an edge $(j, i) \in \gcge$ if and
  only if $x_j$ Granger-causes $x_i$ (with respect to $x$).  That is,
  $$(j, i) \in \gcge \iff j \in \pa{i} \iff x_j \gc x_i.$$
\end{definition}

The edges of the Granger causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with the
\textit{strictly causal} LTI filter $\B_{ij}(z)$ (see Equation
\eqref{eqn:filter_action}).  Thence, the matrix $\B(z)$ is analogous
to a \textit{weighted adjacency matrix}\footnote{\footnotesize We are using the
  convention that $\B_{ij}(z)$ is a filter with input $x_j$ and output
  $x_i$ so as to write the action of the system as $\B(z)x(t)$ with
  $x(t)$ as a column vector.  This competes with the usual convention
  for adjacency matrices where $A_{ij} = 1$ if there is an edge
  $(i, j)$.  In our case, the sparsity pattern of $\B_{ij}$ is the
  \textit{transposed} conventional adjacency matrix.} for the graph $\gcg$.  And,
in the same way that the $k^{\text{th}}$ power of an adjacency matrix
counts the number of paths of length $k$ between nodes,
$(\B(z)^k)_{ij}$ is a filter isolating the ``action'' of $j$ on $i$ at
a time lag of $k$ steps, this is exemplified in the inversion formula
\eqref{eqn:lsi_inversion}.

@@

\subsection{Pairwise Granger Causality}
\label{sec:pwgc}
Recall that Granger causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed, completely
explains the connection from $x_j$ to $x_i$.  In this section we study
\textit{pairwise} Granger causality, and seek to understand when
knowledge of pairwise relations is sufficient to deduce the true fully
conditional relations of $\gcg$.

\begin{definition}[Pairwise Granger causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_j$ Granger-causes $x_i$ with respect only to
  $(x_i, x_j)$.
\end{definition}

@@

\begin{proposition}
  Consider distinct nodes $i, j$ in a Granger causality graph
  $\gcg$.  If

@@

\begin{proposition}
  \label{prop:ancestor_properties}
  If in a Granger causality graph $\gcg$ where $j \pwgc i$ then
  $j \in \anc{i}$ or $\exists k \in \anc{i} \cap\anc{j}$ which is a
  confounder of $(i, j)$.
\end{proposition}

@@

\begin{definition}[Strongly Causal]
  \label{def:strongly_causal}
  We will say that a Granger causality graph $\gcg$ is
  \textit{strongly causal} if there is at most one directed path between
  any two nodes.  Strongly Causal Graphs will be referred to as SCGs.
\end{definition}

@@

\begin{definition}[Persistent]
  We will say that the process $x(t)$ with Granger causality graph
  $\gcg$ is \textit{persistent} if for every $i \in [n]$ and every
  $k \in \anc{i}$ we have $\tau_0(\A_{ik}) < \infty$ and $\tau_\infty(\A_{ik}) = \infty$.
\end{definition}

\begin{remark}
  In the context of Granger causality, ``most'' systems should be
  persistent.  In particular, $\mathsf{\VAR}(p)$ models are likely to
  be persistent since these naturally result in an equivalent
  $\mathsf{MA}(\infty)$ representation, see Example
  \ref{ex:persistent_system}.

@@

\begin{theorem}[Pairwise Recovery]
  \label{thm:scg_recovery}
  If the Granger causality graph $\gcg$ for the process $x(t)$ is a
  strongly causal DAG and Assumption \ref{ass:T_causality} holds, then
  $\gcg$ can be inferred from pairwise causality tests.  The procedure
  can be carried out, assuming we have an oracle for pairwise
  causality, via Algorithm (\ref{alg:pwgr}).
\end{theorem}

@@

  % \BlankLine
  \caption{Pairwise Granger Causality Algorithm}
  \label{alg:pwgr}
  \TitleOfAlgo{Pairwise Graph Recovery}
  \Input{Pairwise Granger causality relations between a persistent
  process of dimension $n$ whose joint Granger causality
  relations are known to form a strongly causal DAG $\gcg$.}
  % \Input{Pairwise Granger causality relations}
  \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
    the graph $\gcg$.}
  \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
    $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
    % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
    $k = 1$ \texttt{\# a counter used only for notation}}
  \BlankLine
  $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
  $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parent-less nodes}\\
  \While{$S_{k - 1} \ne \emptyset$}{
    $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
    $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children}\\

@@

%   contains a loop but is pairwise recoverable, though not by algorithm
%   (\ref{alg:pwgr}).  Clearly, this example is somewhat artificial
%   since when $n = 2$ there is no difference between pairwise
%   Granger causality and joint Granger causality amongst all series --
%   however, one can add any number of nodes having no parents or
%   children to a graph containing a length 2 cycle, in which case the
%   graph clearly remains pairwise recoverable.
% \end{example}

@@

\begin{remark}[MCC as a Support Recovery Measurement]
  We apply Matthew's Correlation Coefficient (MCC)
  \cite{matthews1975comparison} as a statistic for measuring support
  recovery performance (see also \cite{chicco2017ten} tip \# 8).  This
  statistic synthesizes the confusion matrix into a single score
  appropriate for unbalanced labels and is calibrated to fall into the
  range $[-1, 1]$ with $1$ being perfect performance, $0$ being the
  performance of random guessing, and $-1$ being perfectly opposed.
\end{remark}

@@

\section{Conclusion}
\label{sec:conclusion}
In this paper we have argued that considering particular topological
properties of Granger causality networks can provide substantial
insights into the structure of causality graphs with potential for
providing improvements to causality graph estimation when structural
assumptions are met.  In particular, the notion of a strongly-causal
graph has been exploited to establish conditions under which pairwise
causality testing alone is sufficient for recovering a complete
Granger causality graph.  Moreover, examples from the literature
suggest that such topological assumptions may be reasonable in some
applications.  And secondly, even when the strong-causality assumption
is not met, we have provided simulation evidence to suggest that our
pairwise testing algorithm PWGC can still outperform the LASSO and
adaLASSO, both of which are commonly employed in applications.

We emphasize that the causality graph topology is one of the key
defining features of time series analysis in comparison to standard
multivariate regression and therefore advocate for further study of
how different topological assumptions may impact the recovery of
causality graphs.  For example, are there provable guarantees on the
error rate of PWGC when applied to non strongly-causal graphs?  Can
constraint systems or cunning adaptive weighting schemes impose useful
prior knowledge about graph topology for the LASSO algorithm?
Finally, the work of \cite{barnett2015granger} has established the
superiority of Granger causality testing by state space models (as
opposed to pure autoregressions) in many cases.  Combining this work
with our PWGC algorithm (by modifying the approach described in
Section \ref{sec:pairwise_hypothesis_testing} to instead utilize
state-space Granger causality testing) therefore is likely to enable
application to very large networks of time series data which are not
well approximated by finite $\VAR(p)$ models.  Moreover, our
heuristics are in principle applicable in a model-free context.  As
long as a primitive for testing the pairwise causation between two
components of a multivariate time series is available, our methods may
be useful.
